Flaring star-planet interactions in star-planet systems in Kepler and TESS

### 2023-05-08 src/tex/ms.tex
[showyourwork] overleaf sync

    Out of 1811 searched systems, 26 are single star systems that show three or more flares each. For these, we quantified the intensity of excess flaring, and compared it against the expected amplitude of interaction based on the system's parameters. Most systems are neither expected to interact, nor do they show excess flaring. Among those systems that have high expected power of interaction, we see two branches. An inactive one, where no signs of excess flaring appear regardless of favorable conditions, and an active one where expected power and excess flaring are correlated. While this trend is tentative, we also found that the two branches form two distinct populations in spin-orbit commensurability. The active branch is closer to low-integer ratios between rotational and orbital period than the inactive branch. While viewing geometry and intermittency of interaction remain possible explanations, spin-orbit resonances hint at a physical link worth exploring to better understand the underlying mechanisms of star-planet interactions.
    %magnetic field strength (using $Ro$-$B$ relations), orbital parameters, and planet radius.
\end{abstract}

@@

We cannot tell whether an additional individual flare is produced by tidal or magnetic interaction, but as with magnetic interaction, we can look for deviations from a random distribution of flare times in phase with the relevant period. In this case, the relevant period is $P_{orb}/2$.

We applied the same technique as described in Section~\ref{sec:methods:adtest} to determine the significance of a local tidal SPI signal. Following~\citet{ilic2022tidal}, we compare the resulting $p$-values with three different models of tidal interaction in Fig.~\ref{fig:tidal} and list the derived values in Table~(XXXX). The first model measures the torque $\partial L / \partial t$, exerted by angular momentum transfer from star to planet or from planet to star~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$, but our results do not depend on the choice of $Q_*$. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects. \textcolor{red}{Why doesn't the torque depend on the stellar rotation velocity, except for the sign of interaction?}. The second model considers the tidal dissipation timescale -- the shorter the timescale, the higher the power of the interaction~\citep{albrecht2012obliquities}. The third model uses the gravitational perturbation as a proxy for the interaction~\citep{cuntz2000stellar}.

We find that no single system shows significant signs of tidal star-planet interaction. Overall, the signal in phase with $P_{orb}/2$ is smaller than with $P_{orb}$. However, in all three scenarios, the deviation from random flare timing increases with expected tidal interaction strength.

Among the systems, HIP 67522 appears with tentative signs of both tidal and magnetic interaction (measured and expected in the different scenarios). KOI-12, which lacks signal with $P_{orb}$ despite high expected power of magnetic interaction, shows elevated expected and measured tidal interaction. This could be interpreted as KOI-12 b orbiting outside the Alfv\'en radius, unable to interact magnetically, but tidally. We note that KOI may be evolving off the main sequence, so that the tidal interaction estimates might not be accurate for this star. Finally, Kepler-42 is neither expected to interact magnetically, nor did it show any deviation from random flare timing in phase with $P_{orb}$. However, since it is the system with the shortest planetary orbit in our sample, it is both expected to and does show marginal signal of tidal interaction. Kepler-42 is very similar to TOI-540, which also is a small M dwarf with a likely terrestrial planet in a very short orbit. The key difference may be that Kepler-42 is rotating at about 70 days, and therefore expected to have a relatively weak magnetic field, whereas TOI-540 is a young star with a rotation period $<1\,$d. TOI-540 is expected to have lower tidal interaction strength than Kepler-42 due to TOI-540 b's wider orbit, and also shows a weaker tidal SPI signal in all three scenarios.

The results here, as with magnetic SPI, are suggestive, as they are consistent with theoretical expectation, albeit at low significance. Further observations, underway with TESS and soon PLATO, may confirm or reject the trends seen for both types of interaction. As multiple observations can be combined regardless of the gaps in time between individual campaigns with our method, we will become able to sensitively constrain the magnitude of interaction, and the relative contributions from tidal and magnetic SPI.

>>>>

    Out of 1811 searched systems, 26 are single star systems that show three or more flares each. For these, we quantified the intensity of excess flaring, and compared it against the expected amplitude of interaction based on the system's parameters. Most systems are neither expected to interact, nor do they show excess flaring. Among those systems that have high expected power of magnetic interaction, we see two branches. An inactive one, where no signs of excess flaring appear regardless of favorable conditions, and an active one where expected power and excess flaring are correlated. While this trend is tentative, it is consistent across all investigated models for magnetic interaction. Finally, we searched for signs of tidal interaction, that is, excess flaring in phase with the tidal bulges ($P_{orb}/2$), and found a similar trend with different expected powers of tidal interaction. Our results support the view that young systems with close in planets are likely to interact both tidally and magnetically, whereas star-planet interactions in old, slowly rotating systems lack the extended magnetosphere, and interact predominantly tidally.
    %magnetic field strength (using $Ro$-$B$ relations), orbital parameters, and planet radius.
\end{abstract}

@@

We cannot tell whether an additional individual flare is produced by tidal or magnetic interaction, but as with magnetic interaction, we can look for deviations from a random distribution of flare times in phase with the relevant period. In the case of tidal interactions, the relevant period is $P_{orb}/2$.

We applied the same technique as described in Section~\ref{sec:methods:adtest} to determine the significance of a local tidal SPI signal. Following~\citet{ilic2022tidal}, we compare the resulting $p$-values with three different models of tidal interaction in Fig.~\ref{fig:tidal} and list the derived values in Table~(XXXX). The first model measures the torque $\partial L / \partial t$, exerted by angular momentum transfer from star to planet or from planet to star~(\citealt{penev2012constraining}, their Eqns. 1 and 2). We choose the tidal quality factor $Q_*=10^7$, but our results do not depend on the choice of $Q_*$. We also make the simplifying assumption that the rotation period of the convective envelope is the same as the surface rotation period measured as described in Section~\ref{sec:data:rotationperiods}, ignoring differential rotation effects. \textcolor{red}{Why doesn't the torque depend on the stellar rotation velocity, except for the sign of interaction?}. The second model considers the tidal dissipation timescale -- the shorter the timescale, the higher the power of the interaction~\citep{zahn1977tidal}. Following~\citet{ilic2022tidal}, we adopt Eqn.2 from~\citep{albrecht2012obliquities} for the tidal dissipation timescale in the convective envelope. We assume that all stars in our sample have a convective envelope because they flare, and flaring requires a dynamo-generated magnetic field, which can only operate in stars with convective envelopes. Most of our stars are clearly below the Kraft break at spectral type F5~(XXXX), above which stars have a radiative envelope, except for KOI-12, which is an F5 sub-giant~\citep{frasca2016activity}. The third model uses the gravitational perturbation as a proxy for the interaction~\citep{cuntz2000stellar}. All models require the planetary and stellar mass to be known, which we collect from the literature. For missing planetary masses, radii are known, so we can use the mass-radius relations in~\citet{chen2017probabilistic}. We list the masses in Table~(XXXX).

We find that no single system shows significant signs of tidal star-planet interaction. Overall, the signal in phase with $P_{orb}/2$ is smaller than with $P_{orb}$. However, in all three scenarios, the deviation from random flare timing increases with expected tidal interaction strength. We can now compare the magnetic interaction measurements to the tidal ones in some of the conspicuous systems:

Among the systems, HIP 67522 appears with tentative signs of both tidal and magnetic interaction (measured and expected in the different scenarios). On the one hand, since the flares in HIP 67522 mostly appear close to the transit phase, and not during eclipse, this may favor magnetic interaction, and the tidal signal would only be an artifact. On the other hand, the narrow range in which most of the flares on HIP 67522 occurred favor a well-located tidal bulge. It is conceivable that both effects play a role at the same time.

KOI-12, which lacks signal with $P_{orb}$ despite high expected power of magnetic interaction, shows elevated expected and measured tidal interaction. This could be interpreted as KOI-12 b orbiting outside the Alfv\'en radius, unable to interact magnetically, but tidally. We note that KOI may be evolving off the main sequence, so that the tidal interaction estimates might not be accurate for this star. TAP 26, which shows no signs of magnetic SPI, does not show any sign of tidal interaction despite both being high as expected from models. This could again be due to the inclination of TAP 26's orbit, but only if the flares caused by the tidal bulge are not too close to the equator. In that case, TAP 26's tidally induced flares should be more modulated than the magnetically induced ones.

Kepler-42 is neither expected to interact magnetically, nor did it show any deviation from random flare timing in phase with $P_{orb}$. However, since it is the system with the shortest planetary orbit in our sample, it is both expected to and does show marginal signal of tidal interaction. Kepler-42 is very similar to TOI-540, which also is a small M dwarf with a likely terrestrial planet in a very short orbit. The key difference may be that Kepler-42 is rotating at about 70 days, and therefore expected to have a relatively weak magnetic field, whereas TOI-540 is a young star with a rotation period $<1\,$d. TOI-540 is expected to have lower tidal interaction strength than Kepler-42 due to TOI-540 b's wider orbit, and also shows a weaker tidal SPI signal in all three scenarios. K2-25 and AU Mic are similar, but differ from TOI-540 -- both show tentative signs of magnetic SPI, but are consistent within $1\sigma$ with no observable tidal interaction.

%The results here, as with magnetic SPI, are suggestive, as they are consistent with theoretical expectation, albeit at low significance. Further observations, underway with TESS and soon PLATO, may confirm or reject the trends seen for both types of interaction. As multiple observations can be combined regardless of the gaps in time between individual campaigns with our method, we will become able to sensitively constrain the magnitude of interaction, and the relative contributions from tidal and magnetic SPI.
